Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-10-22 manuscript.qmd
add newline to manuscript end

To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023]. It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subsite, lateralization of the primary tumor, and HPV p16 status. Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview. The data from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only consist of patients who all received a neck dissection. The majority of patients from the University Hospital Zürich (USZ) and the Vall d'Hebron Barcelona Hospital (VBH) was treated with definitive radiotherapy. Since sugical treatment is more common for early T-category patients, the ISB and CLB datasets contain a larger portion of early T-category patients compared to the USZ and VBH dataset. For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.

@@

The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023. The Data-in-Brief publications contain a detailed description of the datasets and the data format. These publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital, which will be described in a future publication.

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in blue (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V are healthy). These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}

@@

# Results: Prediction of Risk for Occult Disease

@@

## Limitations and Future Work

>>>>

To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023]. It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subsite, lateralization of the primary tumor, and HPV p16 status. Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview. The data from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only consist of patients who all received a neck dissection. The majority of patients from the University Hospital Zürich (USZ) and the Hospital Vall d'Hebron (HVH) was treated with definitive radiotherapy. Since sugical treatment is more common for early T-category patients, the ISB and CLB datasets contain a larger portion of early T-category patients compared to the USZ and HVH dataset. For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.

@@

The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023. The Data-in-Brief publications contain a detailed description of the datasets and the data format. These publications do not include the most recent dataset addition from HVH, which will be described in a future publication.

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in blue (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V are healthy). These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures
.svg){#fig-model-prevalences-with-ipsi}

@@

# Results: Prediction of Risk for Occult Disease {#sec-results-risk}

@@

In essence, this is a summary of the results in @sec-results-risks. However, it should be considered in light of the limitations outlined in @sec-discussion-limitations below.

## Limitations and Future Work {#sec-discussion-limitations}
